---
title: "Session 3"
author: "Callin Switzer"
date: "September 29, 2015"
output: html_document
---

Topics:

1. Manipulate your data (sub setting with logical statements, cleaning data, making calculations with a data frame) 
2. Export data 
3. Formatting data and visualization using ggplot()
4. ** Loops and apply functions()
5. ** Writing your own functions

**if time


```{r, echo = FALSE}
# print system time, so we know when the last update was
paste("Last knit to html at: ", Sys.time())
```

---

# Review
#### What we learned last time:
1. Set and get working directory
2. Import different file types into R
3. Navigate your data frames, using square-bracket and "$" notation.
4. Quickly visualize your data with the plot(), hist(), and rug() functions and/or car::scatterplotMatrix()

---

```{r}
# sets your working directory
setwd("~/Dropbox/Harvard/RByCall/RClassMaterials/Session3")

# read in pollen dataset
flr <- read.csv("datasets/CallinPollen.csv")
```

---

<span style="color:red; font-family:Georgia; font-size:2em;">
Exercises(0): 
</span>

### 0.1 Set your working directory

### 0.2 Import the "CallinPollen.csv" dataset, and call it "flr".

---


## New Material

---

## 1. Manipulate your data (sub setting with logical statements, cleaning data, making calculations with a data frame)
#### List of logical operators
* "==" equals
* "!=" not equal
* "&" and
* "|" or
* ">" greather
* "<" less
* ">=" greather or equal
* "<=" less or equal
* "%in%" is in  

```{r}
# select rows, all columns
FewRows <- flr[10:20, ]
FewRows

# use logical statements to return logical vectors
amb <- flr$trt == "Ambient" # returns a logical vector that is TRUE when flr$trt is ambient
head(amb)
summary(amb)

# Use "&" to refine further
ambHighHum <- flr$trt == "Ambient" & flr$humidity > 50
summary(ambHighHum)

# Note: can't use x == NA; us is.na() instead
head(flr$openTime == NA) # doesn't work
naOpen <- is.na(flr$openTime)
naOpen
```

---

<span style="color:red; font-family:Georgia; font-size:2em;">
Exercises(1): 
</span>

### 1.1 Make a new logical vector, called "seven" that is TRUE if flr$distance is equal to 7.  Use summary(seven) to check to make sure you have 12 TRUE. 


```{r}
seven <- flr$distance == 7
summary(seven)

```

### 1.2 Make a new logical vector, called "NotAmbHot", where "flr $ temp > 24.5" and flr $ trt is not "Ambient". Use summary(NotAmbHot) to make sure you have 150 TRUE.

```{r}
NotAmbHot <- flr$temp > 24.5 & flr$trt != "Ambient"
summary(NotAmbHot)

```


---



#### We can use logical statements to subset dataframes and clean data
Use square bracket notation

```{r}
# select rows where daysOpen == 1, all columns
DO1 <- flr$daysOpen == 1
d1 <- flr[DO1, ]
summary(d1)

# we can do it all in one step
d1 <- flr[flr$daysOpen == 1, ]
summary(d1)

# select rows where daysOpen does not = 1, and distance == 1
d1d1 <- flr[flr$daysOpen != 1 & flr$distance == 1, ]
head(d1d1)


# cleaning data
hist(flr$slope); rug(flr$slope) # look at histogram and rug

# say we want to drop the points where slope > 0.12
sub1 <- flr[flr$slope < 0.12, ]
hist(sub1$slope); rug(sub1$slope) # check our work

```

#### Calculations within data frames
And tabulating some data from the data frames

```{r}
# say we want to calculate a new measurement that is a combination of slope + int
flr$NewStat <- flr$slope + flr$int
head(flr$NewStat)

# or this will work
flr$NewStat1 <- with(flr, slope * int)
head(flr$NewStat1)

# say we want to change one of the variable types in our data frame
class(flr$daysOpen)
flr$daysOpen <- as.factor(flr$daysOpen)
class(flr$daysOpen)

# or this will work
class(flr$flowerNum)
flr <- within(flr, flowerNum <- as.factor(flowerNum))
class(flr$flowerNum)


# calculate averages by individual
# say we want average glycogen for each individual
URL <- "http://nature.berkeley.edu/~casterln/crawley/rats.txt"
rats <- read.csv(URL, sep = "\t")

summary(rats)
head(rats)
tapply(X = rats$Glycogen, FUN = mean, INDEX = rats$Rat)

# calcuate mean by treatment and individual
tapply(X = rats$Glycogen, FUN = mean, INDEX = interaction(rats$Rat, rats$Treatment))


tab1 <- tapply(X = flr$slope, INDEX = flr$trt, FUN = mean)
class(tab1)
mode(tab1)

# tabulate your data (get counts)
xt1 <- xtabs(formula = ~trt, data = flr)
xt1

# quickly plot the table
barplot(xt1)

# convert to data frame for easy saving
tab1 <- as.data.frame(tab1)
```

---

<span style="color:red; font-family:Georgia; font-size:2em;">
Exercises(1, cont'): 
</span>

### 1.3 Read in the "loblolly.txt" data. Call it "lob".  Make sure header = TRUE, if you use the read.table() function
```{r}
file.show("datasets/loblolly.txt")
lob <- read.table("datasets/loblolly.txt", header = TRUE, sep = " ")
summary(lob)

lob <- read.csv("datasets/loblolly.txt", sep = " ",  header = TRUE)
summary(lob)

```

### 1.4 Make a new smaller dataframe, called lob.sm, where "Seed" is not "311", and where "height" is greater than 5.  Plot a histogram of "height" to prove that there are none less than 5 (feel free to add a rug). 

* check to see if 311 is in your new dataframe with this command: "311 %in% lob.sm$Seed"

     
```{r}
lob.sm <- lob[lob$Seed != 311 & lob$height > 5, ]
hist(lob.sm$height); rug(lob.sm$height)

311 %in% lob.sm$Seed # not there
305 %in% lob.sm$Seed # is there

```


### 1.5 Read in the "yarn.txt" file; call it yarn. Make two separate histograms -- one for the breaks in wool A and one for the breaks in wool B.  
     
*Advanced* plot the histograms side-by-side on one panel, and make sure the x-axis limits are the same for each.  

```{r}
file.show("datasets/yarn.txt")
yarn <- read.table("datasets/yarn.txt", sep = "\t", header = TRUE)

par(mfrow = c(1,2))
limx <- c(0, 80)
hist(yarn$breaks[yarn$wool == "A"], main = "Wool A", xlim = limx)
hist(yarn$breaks[yarn$wool == "B"], main = "Wool B", xlim = limx)
par(mfrow = c(1,1))

## or this
library(ggplot2)
ggplot(yarn, aes(x = breaks)) + 
     geom_histogram() + 
     facet_wrap( ~ wool) + 
     theme_bw()
```


### 1.6 *Advanced* Calculate the mean, median, and range for "breaks" for each type of wool in the dataset.
```{r}
tapply(yarn$breaks, INDEX = yarn$wool, FUN = function(x) c(mean = mean(x), median = median(x), range = range(x)))
```

---

## 2. Export data
```{r}
# Write text files
# can also use write.csv()
tab1
write.table(x = tab1, file = "Tab1.csv", sep = ",",  col.names = FALSE)
file.remove("Tab1.csv") # remove the file

# saving plots
pdf(file = "plot1.pdf", width = 5, height = 4) # open graphics device
# can also use postscript()
plot(x = flr$humidity, y=flr$temp) # make your plot
dev.off() # close the graphics device

# remove the file from my directory
file.remove("plot1.pdf")

# or just export with the "Export" button 
```

---

<span style="color:red; font-family:Georgia; font-size:2em;">
Exercises(2): 
</span>

### 2.1 Read int the CallinPollen.txt file (if you haven't already), and call it flr. Plot x = humidity, y = int.  Export this plot to a file called "humInt.pdf".  Make the width = 4 and height = 3.

```{r}
pdf("humInt.pdf", width = 4, height = 3)
with(flr, plot(x = humidity, y = int))
dev.off()
file.remove("humInt.pdf") # remove file
```

### 2.2 *Advanced* Make that plot look nicer! Change the points to solid dots.  Remove the box around the plot.  Give it a nice title and rename the axes.  Save it with width = 5, height = 4 as "fancyHumInt.pdf". Note: "int"" is a measurement of pollen released from a flower. (hint: type "?plot" to see help)

```{r}
pdf("fancyHumInt.pdf", width = 5, height = 4)
with(flr, plot(x = jitter(humidity), y = int, 
               ylab = "Pollen (Integration Method)",
               xlab = "Relative Humidity (%)",
               bty = "l", 
               main = "Pollen released vs. Humidity", 
               pch = 20, 
               col = rgb(.9, .5, .3, .9)
               ))
dev.off()
file.remove("fancyHumInt.pdf")

```

### 2.3 Subset the flr dataset to contain only rows where "distance" > 9 & the "distance" variable is not an "NA" (hint: use !is.na()). Call the new data frame flr.dist.  Export this small data frame as "flrDist.csv".

```{r}
flr.dist <- flr[flr$distance > 9 & !is.na(flr$distance), ]
write.csv(flr.dist, file = "flrDist.csv")
file.remove("flrDist.csv") # remove
```

---

## 3 Formatting data and visualization using ggplot()

Review: installing new libraries

use install.packages("..."), but you only ned to use this the first time you use the library.  After that, you can just use library("...")

```{r}
# install.packages("ggplot2") # do this only the first time you use a library
library(ggplot2) # now, this package is ready to use
# help(package = "ggplot2") # gets help for this new package or library


# let's try it
ggplot(flr, aes(x= humidity, y = int)) + 
     geom_point() 
```


---

<span style="color:red; font-family:Georgia; font-size:2em;">
Exercises(3): 
</span>

### 3.1 Load the ggplot2 library (install, if you haven't already).


### 3.2  To make sure it works, type the following:

ggplot(flr, aes(x = temp, y = humidity)) +
     geom_point()
     
     
```{r}
ggplot(flr, aes(x = temp, y = humidity)) +
     geom_point()
```

---



#### Ways to format data for ggplot

```{r}
df1 <- data.frame(person = c("bob", "sue", "jack", "ann"), number1 = c(4,5,6,1), number2 = c(2,4,6,6))
df1

# we want to plot persons' number1 and number2, but we'd need them to be in the same column
# wrong
ggplot(df1) + 
     geom_point(aes(x = person, y = number1), color = "red") + 
     geom_point(aes(x = person, y = number2), color = "blue")

# you can reformat
library(reshape2)
df.melt <- melt(df1, id.vars = "person")
head(df.melt)

ggplot(df.melt, aes(x = person, y = value)) +
     geom_point(aes(color = variable))


## formatting for barcharts
ggplot(df1, aes(x = person, y = number1)) + 
     geom_bar(stat = "identity")

## or reshape the data frame
df.long <- data.frame(person = unlist(sapply(as.character(df1$person), 
                                             FUN = function(x) 
                                                  rep(x, times = df1[df1$person == x, "number1"]))))

head(df.long)

ggplot(df.long, aes(x = person)) + 
     geom_bar( fill = "red", alpha = 0.6)

```


## Making nice plots with ggplot

Example: Functional Ecology

http://www.functionalecology.org/view/0/authorGuideline.html

Figures. Please submit electronic artwork as TIFF files (for half-tones) or non-rasterized EPS files (for vector graphics) if possible. Detailed information on the publisher's digital illustration standards is available at http://authorservices.wiley.com/electronicartworkguidelines.pdf. When uploaded the appropriate file designation should be selected from the options on Manuscript Central.

Please ensure that symbols, labels, etc. are large enough for 50% reduction. Figures should not be boxed and tick marks should be on the inside of the axes. 


```{r}
# example with flower data
flr1 <- flr[flr$shakeNumber == 1, ]

# let's rename the plants to make things more interesting
library(plyr)

flr1$plant

flr1$plant <- revalue(flr1$plant, c("big002" = "Solanum dulcamara", "sml001" = "Solanum dulcamarensis"))

plt1 <- ggplot(flr1, aes(y = slope, x = temp)) + 
     geom_point(aes(shape = plant, col = plant))
     
plt1 + theme_bw()

# install.packages("grid")
library(grid) # need this package for the unit() function

# define custom functional ecology theme for ggplot
functEcolTheme <-   theme_bw() + 
                    theme(panel.grid.major = element_blank(), 
                         panel.grid.minor = element_blank(), 
                         panel.border = element_blank(), 
                         panel.background = element_blank(), 
                         axis.line = element_line(colour = "black"), 
                         axis.ticks.length = unit(x = -0.25, units = "cm"), 
                         axis.ticks.margin=unit(0.5, "cm"), 
                         legend.title = element_blank(), 
                         text = element_text(family = "Times New Roman", size = 14), 
                         axis.text = element_text(size = 14), 
                         legend.text = element_text(size = 14))

plt1 + functEcolTheme 


plt2 <- plt1 + 
     functEcolTheme + 
     labs(x = expression(paste("Temperature (",degree,"C)")), 
          y = "Slope of pollen release") + 
     scale_y_log10() + 
     theme(legend.text = element_text(face = "italic"), 
           legend.position = c(.87, .9)) + 
     geom_smooth(method = "lm", se = FALSE, color = "black", lty = 2, lwd = 1) + 
     scale_colour_manual(values = c("grey40", "grey70"), name = "Plant species") + 
     scale_shape_discrete(name = "Plant species") # make sure both color and shape are the same name

plt2

# doesn't work
# ggsave(filename = "ggplt2.pdf", plot = plt2, width = 8, height = 6)

# make an eps file
ggsave(filename = "FunctEcol.eps", device = cairo_pdf, plot = plt2, width = 3, height = 2)
file.remove("FunctEcol.eps") # remove file


# make a pdf
ggsave(filename = "FunctEcol.pdf", device = cairo_pdf, plot = plt2, width = 8, height = 6)
file.remove("FunctEcol.pdf") # remove file

```


### End of class ##

---

<span style="color:red; font-family:Georgia; font-size:2em;">
Exercises(3, cont'): 
</span>

### 3.3 Use ggplot to plot a scatterplot (geom_point()) of this: 
     x = flr$shakeNumber, y = flr$slope

```{r}
ggplot(flr, aes(x = shakeNumber, y = slope)) + 
     geom_point()

```


### 3.4 Add the functional ecology theme we defined earlier to the plot
```{r}
ggplot(flr, aes(x = shakeNumber, y = slope)) + 
     geom_point() + 
     functEcolTheme
```


### 3.5 Add this to your previous graph, and see what happens:
     + geom_smooth()
     
```{r, warning=FALSE, messsage = FALSE}
ggplot(flr, aes(x = shakeNumber, y = slope)) + 
     geom_point() + 
     functEcolTheme + 
     geom_smooth()

```

---

## Making boxplots

```{r}
plt3 <- ggplot(flr1, aes(x = trt, y = int)) + 
     geom_boxplot()
plt3

plt3 + functEcolTheme + 
     theme(axis.text.x = element_text(angle = 45, hjust = 1))

# changing order of the boxes -- change levels of a factor
trtMeans <- tapply(X = flr1$int, FUN = mean, INDEX = flr1$trt)
trtMeans.sort <- trtMeans[order(trtMeans, decreasing = TRUE)]
     
flr1$trt <- factor(flr1$trt, levels = names(trtMeans.sort))

plt3.ordered <- ggplot(flr1, aes(x = trt, y = int)) + 
     geom_boxplot(fill = 'black', alpha = 0.5) + 
     functEcolTheme + 
     theme(axis.text.x = element_text(angle = 45, hjust = 1), 
           plot.margin = unit(c(2,1,1,1), "cm")) + 
     labs(x = "Treatment", y = "Pollen release (integration method)")

plt3.ordered 
plt3.ordered + geom_point()
plt4 <- plt3.ordered + geom_point(position = position_jitter(width = 0.1, height = 0), alpha = 0.5)
plt4

ggsave("boxpltTrt.pdf", device = cairo_pdf, plot = plt4, width = 8, height = 6)
file.remove("boxpltTrt.pdf")
```

---

<span style="color:red; font-family:Georgia; font-size:2em;">
Exercises(3, cont'): 
</span>



### 3.6 Read in the "loblolly.txt" dataset, if you haven't already.  Call it "lob". Use ggplot() to plot a scatterplot (geom_point()) with x= age, y = height.
```{r}
lob <- read.table("datasets/loblolly.txt", sep = " ", header = TRUE)
head(lob)
ggplot(lob, aes(x = age, y = height)) + 
     geom_point()

```

### 3.7 Using the "lob" dataframe, convert "age" to a factor. 
     
     lob$age <- as.factor(lob$age)

### Now make a boxplot where x = age, and y = height with ggplot()

```{r}
lob$age <- as.factor(lob$age)

ggplot(lob, aes(x = age, y = height)) + 
     geom_boxplot()
```

### 3.8 Using the "lob" dataframe, add points to the boxplot from 3.7. *Advanced change the color of the points, based on the "Seed" variable from the "lob" dataframe (you may want to convert it to a factor first)*

```{r}
seed <- as.factor(lob$Seed)

ggplot(lob, aes(x = age, y = height)) + 
     geom_boxplot() + 
     geom_point(aes(color = seed))
```

### 3.9 Add the Functional Ecology theme to your plot
```{r}
ggplot(lob, aes(x = age, y = height)) + 
     geom_boxplot() + 
     geom_point(aes(color = seed)) + 
     functEcolTheme

```



############################### END ############################################







## Loops
```{r}
# for loops repeat a specific number of times
for(i in 1:10){
     print(i + 100)
}

for(ii in df1$person){
    print(rep(ii, times = df1[df1$person == ii, "number1"]))
}


## while loops repeat until something happens
x <- 1
while(x < 10){
     print(x + 100) 
     x <- x + 1
}

## Apply functions are often faster than loops

# for loop
system.time({

     newVec <- numeric()
     for(ii in 1:50000){
          newVec[ii] <- ii + 100
     }
})
head(newVec)


## sapply function
system.time({
     newVec.sa <- sapply(1:100000, FUN = function(x) x + 100)
})

## you can speed up a for-loop by pre-allocating space
system.time({

     newVec <- numeric(100000)
     for(ii in 1:100000){
          newVec[ii] <- ii + 100
     }
})


```


## Writing your own functions
```{r}
# functions take in something, do something, then output something

# example 1
printNum.plus5 <- function(num){
     print(num + 5)
}


printNum.plus5(5)
printNum.plus5(1000)


# example 2
plotWithAxisLabels <- function(x1, y1){
     plot(x = x1, y = y1, xlab = "Here is x1", ylab = "Here is y1", 
          bty = "l", pch = 20)
}

plotWithAxisLabels(x1 = flr$distance, y1 = flr$slope)


# example 3 with sapply
# fake data
df1 <- data.frame(person = c("bob", "sue", "jack", "ann"), number1 = c(4,5,6,1), number2 = c(2,4,6,6))	
df1

# want to convert to long format
longerizer1 <- function(name){
     rep(name, times = df1[df1$person == name, "number1"])
}

longerizer1("bob")
df1
longerizer1("sue")

# we can use this function in sapply
system.time({
    listPerson <- sapply(df1$person, longerizer1)
     nms <- unlist(listPerson) 
})

nms




# we could do the same thing with a for loop
system.time({

     names <- as.character(numeric())
     for(ii in df1$person){
          tmp <- rep(ii, length = df1$number1[df1$person == ii])
          names <- append(names, tmp)
     }
     names <- as.factor(names)
     
})


names


```

